---
title: "Introduction to modeldiag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to modeldiag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(modeldiag)
```

## Overview

The `modeldiag` package provides a unified framework for diagnosing statistical models. It supports linear models, generalized linear models (logistic and Poisson regression), and Cox proportional hazards models.

## Linear Models

For linear models, the package tests for:

- Multicollinearity (VIF)
- Heteroscedasticity (Breusch-Pagan test)
- Autocorrelation (Durbin-Watson test)
- Normality of residuals (Shapiro-Wilk test)
- Influential observations (Cook's distance)

### Example

```{r linear-example}
# Fit a linear model
model_lm <- lm(mpg ~ wt + hp + disp, data = mtcars)

# Run diagnostics
diag_lm <- diagnose_model(model_lm)

# View summary
summary(diag_lm)
```

## Logistic Regression

For logistic regression models, the package tests for:

- Multicollinearity (VIF)
- Linearity of logit (Box-Tidwell test)
- Goodness of fit (Hosmer-Lemeshow test)
- Complete/quasi-complete separation
- Influential observations

### Example

```{r logistic-example}
# Fit a logistic regression model
model_glm <- glm(am ~ wt + hp, data = mtcars, family = binomial)

# Run diagnostics
diag_glm <- diagnose_model(model_glm)

# View summary
summary(diag_glm)
```

## Poisson Regression

For Poisson regression models, the package tests for:

- Multicollinearity (VIF)
- Overdispersion
- Zero-inflation
- Residual analysis
- Influential observations

### Example

```{r poisson-example}
# Fit a Poisson regression model
model_pois <- glm(carb ~ wt + hp, data = mtcars, family = poisson)

# Run diagnostics
diag_pois <- diagnose_model(model_pois)

# View summary
summary(diag_pois)
```

## Survival Models

For Cox proportional hazards models, the package tests for:

- Multicollinearity (VIF with warnings)
- Proportional hazards assumption (Schoenfeld residuals)
- Influential observations (dfbetas)
- Functional form

### Example

```{r survival-example}
library(survival)
data(lung)

# Fit a Cox model
model_cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)

# Run diagnostics
diag_cox <- diagnose_model(model_cox)

# View summary
summary(diag_cox)
```

## Visualization

The package provides comprehensive plotting capabilities:

```{r plotting, fig.height=8, fig.width=8}
# Plot diagnostics for linear model
par(mfrow = c(2, 2))
plot(diag_lm)
```

## Interpreting Results

The `summary()` function provides detailed interpretations of each diagnostic test:

- **PASS**: The assumption is satisfied
- **WARNING**: Potential issues detected that warrant investigation
- **FAIL**: Clear violation of assumptions requiring action

Each test result includes specific guidance on interpretation and potential remedies.