---
title: 'modeldiag: A Unified Framework for Comprehensive Diagnostics of Statistical Models in R'
tags:
  - R
  - statistics
  - linear regression
  - logistic regression
  - Poisson regression
  - survival analysis
  - model diagnostics
authors:
  - name: Emmanuel Adewuyi
    orcid: 0000-0002-7920-6866
    affiliation: 1
  - name: Adewale Lukman
    orcid: 0000-0003-2881-1297
    affiliation: 2
  - name: Abiola Owolabi
    orcid: 0000-0001-9843-5085
    affiliation: 3
affiliations:
  - name: London School of Hygiene and Tropical Medicine, United Kingdom
    index: 1
  - name: University of North Dakota, United States of America
    index: 2
  - name: Ladoke Akintola University of Technology, Nigeria
    index: 3
date: 2 June 2026
bibliography: paper.bib
output: rticles::joss_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 5
)
```

# Summary

Rigorous statistical inference depends on the extent to which a fitted model satisfies its
underlying assumptions. Violations—whether of linearity, homoscedasticity, independence of
errors, distributional form, or proportional hazards—invalidate standard errors, inflate type I
error rates, and distort effect estimates. Despite a rich ecosystem of individual diagnostic
tests in R [@rcore2024], practitioners who work across model families must currently
assemble bespoke workflows from disparate packages such as `lmtest` [@zeileis2002],
`car` [@fox2019], `ResourceSelection` [@lele2019], and `survival` [@therneau2000].

`modeldiag` addresses this fragmentation by providing a single, unified interface for model
diagnostics covering four widely-used model classes: ordinary linear models (`lm`), logistic
regression and Poisson regression (both via `glm`), and Cox proportional hazards models
(`coxph`). A single generic function, `diagnose_model()`, dispatches to the appropriate
battery of checks through R's S3 system, while sixteen individual `check_*()` functions
remain available for targeted investigation. A `summary()` method produces human-readable
interpretations, and a `plot()` method renders publication-quality diagnostic graphics.

# Statement of Need

Applied statisticians, biostatisticians, and data scientists routinely fit models from
multiple families within a single analytical workflow—linear models for continuous outcomes,
logistic regression for binary outcomes, Poisson regression for counts, and Cox models for
time-to-event data. Validating each model currently requires expert knowledge of which
packages provide the relevant test, how arguments differ across packages, and how to
interpret and collate disparate outputs.

Existing tools partially address this need. The `performance` package [@ludecke2021] covers
several assumptions for linear and mixed models and provides a `check_model()` function with
visual output, but it does not support Cox proportional hazards models or offer the full range
of GLM-specific diagnostics (e.g., Hosmer–Lemeshow goodness-of-fit, zero-inflation
detection). The `DHARMa` package [@hartig2022] provides simulation-based residual diagnostics
for GLMs and mixed models, but requires additional expertise in interpreting quantile
residuals and does not handle Cox models. The `car` package [@fox2019] provides generalized
VIF and component-plus-residual plots, but is not designed as a stand-alone diagnostic
workflow.

`modeldiag` fills this gap by:

1. Providing a **consistent S3-dispatch interface** so that a single call to
   `diagnose_model()` applies all relevant tests for the detected model class.
2. Supporting **four model classes** within one package: `lm`, `glm` (binomial and Poisson
   families), and `coxph`.
3. Delivering **interpretable text output** alongside numerical results via `summary()`.
4. Exposing **individual check functions** for reproducible, targeted diagnostics.

The package targets practicing statisticians, epidemiologists, and analysts who need to
validate model assumptions efficiently without switching between multiple packages or
remembering class-specific APIs.

# Statistical Background

## Linear Models

The ordinary linear regression model is written as

$$Y = X\beta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(\mathbf{0},\, \sigma^2 I),$$

where $Y \in \mathbb{R}^n$ is the response vector, $X \in \mathbb{R}^{n \times p}$ is the
design matrix, $\beta \in \mathbb{R}^p$ is the coefficient vector, and the errors $\varepsilon$
are assumed to be independent, homoscedastic, and normally distributed.
Valid inference under this model rests on the following assumptions.

### Multicollinearity

When columns of $X$ are nearly linearly dependent, the ordinary least-squares estimator
$\hat{\beta} = (X^\top X)^{-1} X^\top Y$ is numerically unstable and its variance is
inflated. The Variance Inflation Factor (VIF) for predictor $X_j$ is

$$\mathrm{VIF}_j = \frac{1}{1 - R_j^2},$$

where $R_j^2$ is the coefficient of determination from regressing $X_j$ on all remaining
predictors [@fox1992]. A $\mathrm{VIF}_j \geq 10$ is conventionally treated as indicative of
severe collinearity [@fox2019]. `modeldiag` classifies severity as: negligible ($< 2$),
moderate ($2$–$5$), high ($5$–$10$), and severe ($\geq 10$).

### Heteroscedasticity

Homoscedasticity requires $\mathrm{Var}(\varepsilon_i) = \sigma^2$ for all $i$. When this
assumption fails,

$$\mathrm{Var}(\varepsilon_i) = \sigma_i^2 \neq \sigma^2,$$

ordinary least squares remains unbiased but is no longer the minimum-variance linear unbiased
estimator, and conventional standard errors are incorrect. The Breusch–Pagan test
[@breusch1979] regresses squared residuals on the original predictors; the test statistic is
asymptotically $\chi^2$ under homoscedasticity. `modeldiag` implements this test via
`lmtest::bptest()`.

### Autocorrelation

Independence of errors requires $\mathrm{Cov}(\varepsilon_i, \varepsilon_j) = 0$ for $i \neq
j$. First-order autocorrelation, $\varepsilon_i = \rho \varepsilon_{i-1} + \nu_i$, commonly
arises in time-ordered data and inflates apparent precision. The Durbin–Watson statistic
[@durbin1950; @durbin1951] is

$$DW = \frac{\sum_{i=2}^{n}(\hat{\varepsilon}_i - \hat{\varepsilon}_{i-1})^2}{\sum_{i=1}^{n}\hat{\varepsilon}_i^2},$$

with values close to 2 indicating no autocorrelation and values near 0 or 4 indicating
positive or negative autocorrelation respectively. `modeldiag` implements this via
`lmtest::dwtest()`.

### Linearity of the Conditional Mean

The classical Ramsey RESET test [@ramsey1969] evaluates whether the conditional mean
$\mathbb{E}[Y \mid X]$ is correctly specified as $X\beta$ by augmenting the model with
powers of the fitted values $\hat{Y}^2, \hat{Y}^3, \ldots$ and testing their joint
significance. Rejection indicates that non-linear transformations of the predictors may be
required.

### Normality of Errors

Small-sample inference ($t$-tests, $F$-tests) assumes $\varepsilon \sim \mathcal{N}(0,
\sigma^2)$. The Shapiro–Wilk test [@shapiro1965] evaluates this assumption on the
residual vector $\hat{\varepsilon}$. The test statistic is

$$W = \frac{\left(\sum_{i=1}^{n} a_i \hat{\varepsilon}_{(i)}\right)^2}{\sum_{i=1}^{n}(\hat{\varepsilon}_i - \bar{\varepsilon})^2},$$

where $\hat{\varepsilon}_{(i)}$ denotes the $i$-th order statistic and $\mathbf{a}$ is a
vector of coefficients derived from the covariance matrix of normal order statistics. Values
of $W$ close to 1 support normality.

### Influential Observations

Cook's distance [@cook1977; @cook1982] measures the aggregate shift in $\hat{\beta}$ when
observation $i$ is deleted:

$$D_i = \frac{\left(\hat{\beta} - \hat{\beta}_{(i)}\right)^\top X^\top X \left(\hat{\beta} - \hat{\beta}_{(i)}\right)}{p\,\hat{\sigma}^2},$$

where $\hat{\beta}_{(i)}$ is the estimate computed without the $i$-th observation, and $p$
is the number of parameters. A conventional threshold of $D_i > 4/n$ flags potentially
influential observations.

## Logistic Regression

Logistic regression models a binary outcome $Y \in \{0, 1\}$ through

$$\mathrm{logit}\!\left(P(Y = 1 \mid X)\right) = \log\!\left(\frac{p}{1-p}\right) = X\beta,$$

where $p = P(Y = 1 \mid X)$. Key assumptions concern the functional form of the link, model
calibration, and numerical stability of the estimates.

### Linearity of the Logit

The model requires a linear relationship between each continuous predictor and the
log-odds. The Box–Tidwell transformation [@box1962] tests this by augmenting the model with
interaction terms $X_j \ln(X_j)$ and evaluating whether their coefficients depart
significantly from zero. A significant result suggests that $X_j$ enters the logit non-linearly
and that a transformation may be necessary.

### Goodness of Fit: Hosmer–Lemeshow Test

The Hosmer–Lemeshow test [@hosmer2013] partitions observations into $g$ groups of
approximately equal size based on estimated probabilities, then computes a $\chi^2$ statistic
comparing observed event counts $O_k$ to expected counts $E_k$ within each group:

$$\hat{C} = \sum_{k=1}^{g} \frac{(O_k - E_k)^2}{E_k(1 - E_k/n_k)}.$$

Under good fit $\hat{C} \sim \chi^2_{g-2}$; a small $p$-value indicates lack of fit. The
number of groups is set to $g = \min(10,\, \lfloor n/5 \rfloor)$ to maintain adequate cell
counts.

### Complete and Quasi-complete Separation

Separation occurs when a linear combination of the predictors perfectly predicts the binary
outcome. Maximum-likelihood estimates do not exist in finite form (logistic regression
coefficients diverge to $\pm\infty$), and standard errors become unreliable. `modeldiag`
detects separation by inspecting whether any estimated coefficients are infinite or `NA` and
whether the fitting algorithm converged.

## Poisson Regression

Poisson regression models a non-negative integer count outcome $Y$ as

$$Y \mid X \sim \mathrm{Poisson}(\lambda), \quad \log(\lambda) = X\beta.$$

The Poisson distribution has the equidispersion property $\mathrm{Var}(Y) = \mathbb{E}(Y) = \lambda$.

### Overdispersion

Overdispersion occurs when the observed variance exceeds the mean:

$$\mathrm{Var}(Y) > \mathbb{E}(Y).$$

This violates the Poisson equidispersion assumption, causes underestimated standard errors,
and inflates type I error rates. A practical diagnostic is the ratio of the residual deviance
to its degrees of freedom [@dean1989; @cameron2013]:

$$\phi = \frac{D_{\text{resid}}}{\text{df}_{\text{resid}}}.$$

Values of $\phi$ substantially greater than 1 indicate overdispersion; `modeldiag` flags
$\phi > 1.5$ as evidence of overdispersion. In practice, overdispersion is addressed by
switching to a negative binomial model or a quasi-Poisson specification.

### Zero Inflation

Zero inflation arises when the observed proportion of zeros exceeds what is expected under a
Poisson model with the estimated mean. If

$$\hat{p}_0^{\,\text{obs}} > 1.5 \times \hat{p}_0^{\,\text{exp}},$$

where $\hat{p}_0^{\,\text{exp}} = \frac{1}{n}\sum_{i=1}^{n} e^{-\hat{\lambda}_i}$ is the
average Poisson probability of a zero, `modeldiag` flags the model as potentially zero-inflated.
This heuristic prompts the analyst to consider zero-inflated Poisson (ZIP) or zero-inflated
negative binomial (ZINB) models.

## Cox Proportional Hazards Model

The Cox model [@cox1972] specifies the hazard for subject $i$ as

$$h(t \mid X_i) = h_0(t)\,\exp(X_i\beta),$$

where $h_0(t)$ is an unspecified baseline hazard function and $\exp(\beta_j)$ is interpreted
as the hazard ratio associated with a one-unit increase in $X_j$.

### Proportional Hazards Assumption

The model assumes that the hazard ratio between any two subjects is constant over time,
i.e., the log hazard ratio $X_i \beta$ does not depend on $t$. Schoenfeld residuals
[@schoenfeld1982] provide a test of this assumption. For covariate $j$, the Schoenfeld
residual at event time $t_k$ is the difference between the covariate value of the subject
who experienced the event and its risk-set-weighted mean. Under the proportional hazards
assumption these residuals should be uncorrelated with time. The test statistic derived by
@grambsch1994 is used to formally test

$$H_0: \frac{\partial \beta(t)}{\partial t} = 0,$$

via a weighted correlation between the scaled Schoenfeld residuals and a monotone function
of time. `modeldiag` implements this via `survival::cox.zph()`.

### Influential Observations in the Cox Model

Influence in the Cox model is assessed via dfbeta residuals, which approximate the change in
each coefficient $\hat{\beta}_j$ upon deletion of observation $i$. Observations for which
$|\mathrm{dfbeta}_{ij}| > 0.2$ for any coefficient $j$ are flagged as potentially
influential.

### Functional Form

Martingale residuals—the difference between the observed event indicator and the expected
cumulative hazard—can be plotted against continuous predictors to detect non-linear
functional form relationships not captured by $X\beta$.

# Package Design

`modeldiag` is implemented as a standard R package with an S3 dispatch system. The generic
`diagnose_model()` dispatches to `diagnose_model.lm()`, `diagnose_model.glm()`, and
`diagnose_model.coxph()` based on the class of the fitted model. Each method executes the
appropriate battery of `check_*()` functions and returns an object of S3 class
`"model_diagnostics"`, a named list containing the model type, the original model object,
and the results of all diagnostic tests. The `print()`, `summary()`, and `plot()` methods
dispatch on this class.

The sixteen exported `check_*()` functions can also be called independently, enabling
targeted diagnostics within reproducible analysis pipelines. The package depends on `car`
[@fox2019], `lmtest` [@zeileis2002], `ResourceSelection` [@lele2019], and `survival`
[@therneau2000] for the underlying test computations.

# Usage

## Linear Models

```{r lm-full}
library(modeldiag)

model_lm <- lm(mpg ~ wt + hp + disp, data = mtcars)
diag_lm  <- diagnose_model(model_lm)
summary(diag_lm)
```

Individual diagnostic checks can be called separately for targeted analysis:

```{r lm-individual}
check_vif(model_lm)
check_heteroskedasticity(model_lm)
check_autocorrelation(model_lm)
check_linearity(model_lm)
check_normality(model_lm)
check_outliers(model_lm)
```

Diagnostic plots for the linear model are obtained with:

```{r lm-plot, fig.cap="Diagnostic plots for the linear model fitted to the mtcars dataset. Panels show residuals versus fitted values, a normal Q-Q plot of residuals, scale-location, residuals versus leverage, and the autocorrelation function of residuals.", fig.height=7}
plot(diag_lm)
```

## Logistic Regression

```{r logistic-full}
model_glm <- glm(am ~ wt + hp, data = mtcars, family = binomial)
diag_glm  <- diagnose_model(model_glm)
summary(diag_glm)
```

Individual logistic regression diagnostics:

```{r logistic-individual}
check_vif(model_glm)
check_box_tidwell(model_glm)
check_hosmer_lemeshow(model_glm)
check_influential_glm(model_glm)
check_separation(model_glm)
```

## Poisson Regression

```{r poisson-full}
model_pois <- glm(carb ~ wt + hp, data = mtcars, family = poisson)
diag_pois  <- diagnose_model(model_pois)
summary(diag_pois)
```

Individual Poisson diagnostics:

```{r poisson-individual}
check_vif(model_pois)
check_overdispersion(model_pois)
check_zero_inflation(model_pois)
check_residual_analysis(model_pois)
```

## Cox Proportional Hazards Model

```{r cox-full}
library(survival)
data(lung)

model_cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)
diag_cox  <- diagnose_model(model_cox)
summary(diag_cox)
```

Individual Cox model diagnostics:

```{r cox-individual}
check_proportional_hazards(model_cox)
check_influential_coxph(model_cox)
check_functional_form_coxph(model_cox)
```

Diagnostic plots for the Cox model display Schoenfeld residual trends, dfbeta influence
plots, and martingale residuals against each continuous predictor:

```{r cox-plot, fig.cap="Diagnostic plots for the Cox proportional hazards model. The Schoenfeld residual plot assesses the proportional hazards assumption over time.", fig.height=5}
plot(diag_cox)
```

# Acknowledgements

The authors thank the maintainers of `car`, `lmtest`, `ResourceSelection`, and `survival`
for the foundational test implementations that `modeldiag` builds upon.

# References
